---
title: "Census of discoverable diversity: Analyses"
author: "Vishnu Prasoodanan, Oleks Maistrenko & Sebastian Schmidt"
date: "2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(stringsAsFactors = FALSE)
```

## Prepare environment

```{r}
library(tidyverse)
library(extrafont)
library(ape)
library(phytools)
library(phangorn)
library(castor)

extrafont::loadfonts()
```

Set paths and parameters.

```{r}
#Preallocate global data structures
PARAM <- list()

#Set relevant folder names
PARAM$folder.Rmd <- paste0(getwd(), "/")
PARAM$folder.base <- gsub("Rmd/", "", PARAM$folder.Rmd)
PARAM$folder.data <- paste0(PARAM$folder.base, "data/")
PARAM$folder.data.rarefaction <- paste0(PARAM$folder.data, "hmm_gtdb.clustering/")
```

Plotting colours.

```{r}
col.pg3 <-   "#1F78B4"
col.gtdb <-   "#A6CEE3"
col.spire.mag <- "#FDBF6F"
col.unbinned.non_singleton <- "#FFE206"
col.unbinned <-  "#ffffb3"
```

## Read data

Load sample metadata.

```{r}
PARAM$file.sample_data <- paste0(PARAM$folder.base, "metadata/data.sample.clean.Rdata")
load(PARAM$file.sample_data)
```

Load HMM data.

```{r}
PARAM$file.hmm <-  paste0(PARAM$folder.data, "dat.hmm.gtdb_207.Rdata")
load(PARAM$file.hmm)
```


Get a list of marker genes that are "redundant" (i.e., used in both the bacterial and archaeal domain).

```{r}
dat.hmm %>%
  group_by(marker_id) %>%
  tally() %>%
  filter(n > 1) %>%
  pull(marker_id) -> drop.redundant.markers
```

Load clustering summary data

```{r}
PARAM$file.clustering.summary <- paste0(PARAM$folder.data, "00_SUMMARY.spire_gtdb.clustered.rarefaction.Rdata")
load(PARAM$file.clustering.summary)
```

Prepare and re-categorize samples.

```{r}
order.by_h5 <- c(
  "human gut (adult, healthy)",
  "human gut (adult, diseased)",
  "human gut (non-industrialized)",
  "human gut (infant)",
  "human gut (child)",
  "human gut (elderly)",
  "pig gut",
  "bovine gut",
  "mammalian gut (other)",
  "rumen",
  "bird gut",
  "animal gut (other)",
  "human oral",
  "human skin",
  "human airways",
  "human urogenital",
  "human (other)",
  "built environment",
  "wastewater",
  "rhizosphere",
  "phyllosphere",
  "plant host (other)",
  "soil",
  "wetland",
  "marine",
  "fresh water",
  "groundwater",
  "hot spring",
  "hydrothermal vent",
  "aquatic (other)",
  "air",
  "other"
)

dat.sample.rarefy <- dat.sample.clean
```

# RED-based analyses

## Load data

Loop through marker genes and load data.

```{r}
collect.convert_counts <- collect.clade_sizes <- tibble()

for (file in list.files(PARAM$folder.data, pattern = "clade_counts_v3.tab")) {
  marker_id <- file %>% str_remove("_convert.+")
  
  writeLines(paste(date(), "=>", marker_id))
  
  #Load clade counts
  collect.convert_counts <- bind_rows(
    collect.convert_counts,
    read_tsv(paste0(PARAM$folder.data, file), show_col_types = F) %>%
      mutate(marker_id = marker_id) %>%
      select(marker_id, phylum, class, order, family, genus)
  )
  
  #Load clade sizes
  collect.clade_sizes <- bind_rows(
    collect.clade_sizes,
    read_tsv(paste0(PARAM$folder.data, marker_id, "_collect.clade_sizes_v3.tab"), show_col_types = F) %>%
      group_by(level, node) %>%
      summarise(
        source = case_when(
          sum(source == "pg3") > 0 ~ "pg3",
          sum(source == "gtdb_r220") > 0 ~ "gtdb_r220",
          sum(source == "spire.mag") > 0 ~ "spire.mag",
          T ~ "spire.unbinned"
        ),
        size = sum(n)
      ) %>%
      ungroup() %>%
      mutate(marker_id = marker_id) %>%
      mutate(level = fct_relevel(level, "phylum", "class", "order", "family", "genus")) %>%
      select(marker_id, level, node, source, size)
  )
}

```

## Analyze data

Get n of tips in each tree.

```{r, fig.height=6}
#Find predicted total phyla for filtering
use.marker_id.arch <- collect.clade_sizes %>%
  filter(level == "phylum") %>%
  group_by(marker_id) %>%
  tally %>%
  mutate(median.count = median(n)) %>%
  filter(n >= 0.5 * median.count & n <= 2 * median.count) %>%
  filter(! marker_id %in% c(
    "TIGR00111",
    "PF01280.21"
  )) %>%
  pull(marker_id)

collect.tree_sizes <- collect.clade_sizes %>%
  filter(marker_id %in% use.marker_id.arch) %>%
  filter(level == "phylum") %>%
  group_by(marker_id) %>%
  summarise(n.tips = sum(size)) %>%
  filter(n.tips <= 2 * median(n.tips) & n.tips >= 0.5 * median(n.tips)) %>%
  mutate(
    conversion.tree_size = n.tips / median(n.tips)
  )

collect.tree_sizes %>%
  arrange(desc(n.tips)) %>%
  mutate(marker_id = fct_relevel(marker_id, marker_id)) %>%
  ggplot(aes(x = n.tips, y = marker_id)) +
  geom_vline(aes(xintercept = mean(n.tips))) +
  geom_vline(aes(xintercept = median(n.tips)), colour = "green") +
  geom_point() +
  theme_minimal()
```

First, go for clade counts only, with and without normalization.

```{r, fig.width=12, fig.height=8}
proc.clade_sizes <- collect.clade_sizes %>%
  filter(marker_id %in% use.marker_id.arch) %>%
  group_by(marker_id, level, source) %>%
  summarise(
    n = n(),
    n.tips = sum(size)
  ) %>%
  ungroup() %>%
  left_join(
    collect.convert_counts %>%
      pivot_longer(phylum:genus, names_to = "level", values_to = "conversion"),
    by = c("marker_id", "level")
  ) %>%
  left_join(collect.tree_sizes, by = "marker_id", suffix = c("", ".full")) %>%
  mutate(
    n.adj = n / conversion,
    n.adj.n_tips = n / conversion.tree_size,
    source = fct_relevel(source, rev(c("pg3", "gtdb_r220", "spire.mag", "spire.unbinned"))),
    level = fct_relevel(level, "phylum", "class", "order", "family", "genus")
  ) %>%
  #Reorder labels within levels
  group_by(marker_id, level) %>%
  mutate(n.tot = sum(n)) %>%
  ungroup() %>%
  mutate(marker_id.reordered = tidytext::reorder_within(marker_id, desc(n.tot), within = level))

proc.clade_sizes %>%
  ggplot(aes(x = n, y = marker_id, fill = source)) +
  geom_bar(stat = "identity") +
  #Add lines indicating GTDB r220 reference clade counts
  geom_vline(
    data = tibble(
      #Clade counts from GTDB r220
      n = c(19, 64, 166, 564, 1847),
      level = c("phylum", "class", "order", "family", "genus")
    ) %>%
       mutate(level = fct_relevel(level, "phylum", "class", "order", "family", "genus")),
    aes(xintercept = n),
    colour = "darkgrey",
    linetype = "dashed"
  ) +
  #Add lines indicating median counts
  geom_vline(
    data = . %>%
      select(marker_id, level, source, n) %>%
      pivot_wider(values_from = n, names_from = source) %>%
      mutate(gtdb_r220 = pg3 + gtdb_r220) %>%
      mutate(spire.mag = gtdb_r220 + spire.mag) %>%
      mutate(spire.unbinned = spire.mag + spire.unbinned) %>%
      pivot_longer(c(pg3, gtdb_r220, spire.mag, spire.unbinned), names_to = "source", values_to = "n") %>%
      group_by(level, source) %>%
      summarise(n = median(n)) %>%
      ungroup() %>%
      mutate(source = fct_relevel(source, rev(c("pg3", "gtdb_r220", "spire.mag", "spire.unbinned")))),
    aes(xintercept = n, colour = source)
  ) +
  xlab("number of predicted clades") +
  ylab("marker gene") +
  scale_color_manual(values = rev(c(col.pg3, col.gtdb, col.spire.mag, col.unbinned.non_singleton))) +
  scale_fill_manual(values = rev(c(col.pg3, col.gtdb, col.spire.mag, col.unbinned.non_singleton))) +
  facet_wrap(. ~ level, scales = "free_x", nrow = 1) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )
```

Explore the habitat "occupancy" of clades at each level.

```{r, fig.width=8, fig.height=6}
tot.clade_sizes <- collect.clade_sizes %>%
  group_by(marker_id, level) %>%
  summarise(n.total = n())

tot.clade_sizes.summarised <- collect.clade_sizes %>%
  group_by(marker_id, source, level) %>%
  summarise(n.total = n()) %>%
  filter(marker_id %in% collect.h5_counts$marker_id) %>%
  group_by(source, level) %>%
  summarise(count = mean(n.total))

tot.clade_sizes.summarised.for_plotting <- tot.clade_sizes.summarised %>%
  pivot_wider(names_from = source, values_from = count) %>%
  mutate(
    pg3__bl = 0,
    gtdb_r220__bl = pg3,
    spire.mag__bl = pg3 + gtdb_r220,
    spire.unbinned__bl = pg3 + gtdb_r220 + spire.mag,
    pg3__ceil = pg3,
    gtdb_r220__ceil = pg3 + gtdb_r220,
    spire.mag__ceil = pg3 + gtdb_r220 + spire.mag,
    spire.unbinned__ceil = pg3 + gtdb_r220 + spire.mag + spire.unbinned
  ) %>%
  select(level, pg3__bl:spire.unbinned__ceil) %>%
  pivot_longer(pg3__bl:spire.unbinned__ceil) %>%
  separate(name, sep = "__", into = c("source", "limit")) %>%
  pivot_wider(names_from = "limit", values_from = "value")

#Make plot
p.clades.by_h5.arch <- collect.h5_counts %>%
  filter(marker_id %in% collect.h5_counts$marker_id) %>%
  left_join(tot.clade_sizes, by = c("marker_id", "level")) %>%
  mutate(prevalence = n / n.total) %>%
  group_by(level, h5, source) %>%
  summarise(
    prevalence = mean(prevalence),
    count = mean(prevalence) * mean(n.total)
  ) %>%
  ungroup() %>%
  left_join(tot.clade_sizes.summarised.for_plotting, by = c("level", "source")) %>%
  mutate(count.adj = count + bl) %>%
  bind_rows(
    tot.clade_sizes.summarised.for_plotting %>%
      mutate(h5 = "all") %>%
      rename(count.adj = ceil)
  ) %>%
  mutate(h5 = fct_relevel(h5, c("all", levels(collect.h5_counts$h5)))) %>%
  mutate(level = fct_relevel(level, "phylum", "class", "order", "family", "genus")) %>%
  mutate(source = fct_relevel(source, rev(c("pg3", "gtdb_r220", "spire.mag", "spire.unbinned")))) %>%
  mutate(
    xmin = as.numeric(h5) - 0.5,
    xmax = as.numeric(h5) + 0.5
  ) %>%
  filter(level %in% c("phylum", "genus")) %>%
  filter(!is.na(h5)) %>%
  #Plot
  ggplot(aes(x = h5, y = count.adj, ymin = bl, group = source, fill = source)) +
  #Add low-opacity background shading rectangles
  geom_rect(
    data = tot.clade_sizes.summarised.for_plotting %>%
      mutate(source = fct_relevel(source, rev(c("pg3", "gtdb_r220", "spire.mag", "spire.unbinned")))) %>%
      filter(level %in% c("phylum", "genus")),
    aes(x = 34, xmin = 0.5, xmax = 33.5, y = ceil, ymin = bl, ymax = ceil, fill = source),
    alpha = 0.1
  ) +
  #Mark borders between sources
  geom_hline(
    data = tot.clade_sizes.summarised.for_plotting %>%
      mutate(source = fct_relevel(source, rev(c("pg3", "gtdb_r220", "spire.mag", "spire.unbinned")))) %>%
      filter(level %in% c("phylum", "genus")),
    aes(yintercept = ceil, colour = source),
    linetype = "dashed"
  ) +
  #Add data series
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymax = count.adj)
  ) +
  #Adjust scales
  scale_fill_manual(values = rev(c(col.pg3, col.gtdb, col.spire.mag, col.unbinned.non_singleton))) +
  scale_colour_manual(values = rev(c(col.pg3, col.gtdb, col.spire.mag, col.unbinned.non_singleton))) +
  scale_x_discrete(guide = guide_axis(angle = -45), position = "top") +
  ylab("number of detected clade-level groups") +
  #scale_size(range = c(0,6)) +
  #Grid by taxonomic resolution
  facet_grid(level ~ ., scales = "free_y") +
  #Prettify
  theme_minimal(base_family = "Avenir Next", base_size = 8) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(linewidth = 0.1, colour = "darkgrey"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(family = "Avenir Next", size = 8, colour = "#252525"),
    axis.text = element_text(family = "Avenir Next", size = 5, colour = "#252525"),
    strip.text = element_text(family = "Avenir Next", size = 8, colour = "#252525")
  ) +
  NULL
p.clades.by_h5.arch

ggsave(
  p.clades.by_h5.arch,
  filename = paste0(PARAM$folder.base, "results/clade_counts.by_h5.pdf"),
  width = 90,
  height = 60,
  unit = "mm",
  device = cairo_pdf
)
```

Repeat for Bacteria.

Get n of tips in each tree.

```{r, fig.height=12}
collect.tree_sizes <- collect.clade_sizes.bac %>%
  filter(level == "phylum") %>%
  group_by(marker_id) %>%
  summarise(n.tips = sum(size)) %>%
  filter(n.tips <= 2 * median(n.tips) & n.tips >= 0.5 * median(n.tips)) %>%
  mutate(
    conversion.tree_size = n.tips / median(n.tips)
  )

collect.tree_sizes %>%
  arrange(desc(n.tips)) %>%
  mutate(marker_id = fct_relevel(marker_id, marker_id)) %>%
  ggplot(aes(x = n.tips, y = marker_id)) +
  geom_vline(aes(xintercept = mean(n.tips))) +
  geom_vline(aes(xintercept = median(n.tips)), colour = "green") +
  geom_point() +
  theme_minimal()
```

First, go for clade counts only, with and without normalization.

```{r, fig.width=12, fig.height=12}
#Find predicted total phyla for filtering
use.marker_id <- collect.clade_sizes.bac %>%
  filter(level == "phylum") %>%
  #filter(source %in% c("pg3", "gtdb_r220")) %>%
  group_by(marker_id) %>%
  tally %>%
  mutate(median.count = median(n)) %>%
  #Filter marker genes (in addition to manual filters based on curation applied upstream)
  filter(n >= 0.5 * median.count & n <= 2 * median.count) %>%
  pull(marker_id)

proc.clade_sizes <- collect.clade_sizes.bac %>%
  filter(marker_id %in% use.marker_id) %>%
  group_by(marker_id, level, source) %>%
  summarise(
    n = n(),
    n.tips = sum(size)
  ) %>%
  ungroup() %>%
  left_join(
    collect.convert_counts %>%
      pivot_longer(phylum:genus, names_to = "level", values_to = "conversion"),
    by = c("marker_id", "level")
  ) %>%
  left_join(collect.tree_sizes, by = "marker_id", suffix = c("", ".full")) %>%
  mutate(
    n.adj = n / conversion,
    n.adj.n_tips = n / conversion.tree_size,
    source = fct_relevel(source, rev(c("pg3", "gtdb_r220", "spire.mag", "spire.unbinned"))),
    level = fct_relevel(level, "phylum", "class", "order", "family", "genus")
  ) %>%
  #Reorder labels within levels
  group_by(marker_id, level) %>%
  mutate(n.tot = sum(n)) %>%
  ungroup() %>%
  mutate(marker_id.reordered = tidytext::reorder_within(marker_id, desc(n.tot), within = level))

proc.clade_sizes %>%
  ggplot(aes(x = n, y = marker_id, fill = source)) +
  geom_bar(stat = "identity") +
  #Add lines indicating GTDB r220 reference clade counts
  geom_vline(
    data = tibble(
      #Clade counts from GTDB r220
      n = c(175, 538, 1840, 4870, 23112),
      level = c("phylum", "class", "order", "family", "genus")
    ) %>%
       mutate(level = fct_relevel(level, "phylum", "class", "order", "family", "genus")),
    aes(xintercept = n),
    colour = "darkgrey",
    linetype = "dashed"
  ) +
  #Add lines indicating median counts
  geom_vline(
    data = . %>%
      select(marker_id, level, source, n) %>%
      pivot_wider(values_from = n, names_from = source) %>%
      mutate(gtdb_r220 = pg3 + gtdb_r220) %>%
      mutate(spire.mag = gtdb_r220 + spire.mag) %>%
      mutate(spire.unbinned = spire.mag + spire.unbinned) %>%
      pivot_longer(c(pg3, gtdb_r220, spire.mag, spire.unbinned), names_to = "source", values_to = "n") %>%
      group_by(level, source) %>%
      summarise(n = median(n)) %>%
      ungroup() %>%
      mutate(source = fct_relevel(source, rev(c("pg3", "gtdb_r220", "spire.mag", "spire.unbinned")))),
    aes(xintercept = n, colour = source)
  ) +
  xlab("number of predicted clades") +
  ylab("marker gene") +
  scale_color_manual(values = rev(c(col.pg3, col.gtdb, col.spire.mag, col.unbinned.non_singleton))) +
  scale_fill_manual(values = rev(c(col.pg3, col.gtdb, col.spire.mag, col.unbinned.non_singleton))) +
  facet_wrap(. ~ level, scales = "free_x", nrow = 1) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )
```

Export table with estimated clade counts.

```{r}
res.clade_counts <- collect.clade_sizes.bac %>%
  filter(marker_id %in% use.marker_id) %>%
  mutate(domain = "Bacteria") %>%
  bind_rows(
    collect.clade_sizes %>%
    filter(marker_id %in% use.marker_id.arch) %>%
    mutate(domain = "Archaea")
  ) %>%
  group_by(domain, marker_id, level, source) %>%
  tally %>%
  ungroup() %>%
  pivot_wider(names_from = "source", values_from = "n", values_fill = 0) %>%
  mutate(ref = pg3 + gtdb_r220) %>%
  select(domain, marker_id, level, ref, spire.mag, spire.unbinned) %>%
  pivot_longer(ref:spire.unbinned, names_to = "source", values_to = "n") %>%
  mutate(
    level = fct_relevel(level, c("phylum", "class", "order", "family", "genus")),
    source = fct_relevel(source, c("ref", "spire.mag", "spire.unbinned"))
  )

tab.clade_counts <- res.clade_counts %>%
  group_by(domain, level, source) %>%
  summarise(
    median = round(median(n, na.rm=T)),
    sd = round(sd(n, na.rm = T))
  ) %>%
  ungroup() %>%
  arrange(domain, level, source) %>%
  mutate(number_of_clades = str_c(median, " ± ", sd)) %>%
  select(domain, level, source, number_of_clades) %>%
  pivot_wider(names_from = "level", values_from = "number_of_clades")
tab.clade_counts
```

Plot estimated clade counts.

```{r}
p.clade_counts <- res.clade_counts %>%
  ggplot(aes(x = source, y = n, fill = source)) +
  #Add reference hlines
  geom_hline(
    data = bind_rows(
      tibble(
        domain = "Archaea",
        n = c(1, 100, 1000)
      ),
      tibble(
        domain = "Bacteria",
        n = c(1, 10, 100, 1000, 10000)
      )
    ),
    aes(yintercept = n),
    colour = "darkgrey",
    linewidth = 0.1
  ) +
  #Add boxplot
  geom_boxplot(aes(group = source), fill = NA, outlier.colour = NA, colour = "#525252", linewidth = 0.4) +
  #Add data series
  geom_point(
    position = position_jitterdodge(),
    shape = 21,
    size = 1.2,
    linewidth = 0.01,
    colour = "#737373",
  ) +
  #Add ref lines for GTDB
  geom_hline(
    data = bind_rows(
      tibble(
        domain = "Archaea",
        level = c("phylum", "class", "order", "family", "genus"),
        n = c(19, 64, 166, 564, 1847)
      ),
      tibble(
        domain = "Bacteria",
        level = c("phylum", "class", "order", "family", "genus"),
        n = c(175, 538, 1840, 4870, 23112)
      )
    ) %>%
      mutate(level = fct_relevel(level, c("phylum", "class", "order", "family", "genus"))),
    aes(yintercept = n),
    colour = col.gtdb,
    linetype = "dashed",
    linewidth = 0.5
  ) +
  scale_fill_manual(values = c(col.gtdb, col.spire.mag, col.unbinned.non_singleton)) +
  scale_y_continuous(
    trans = "log10",
    breaks = rep(1:9, each = 5) * c(1, 10, 100, 1000, 10000)
  ) +
  facet_grid(
    domain ~ level,
    scales = "free_y",
    switch = "both",
    #nrow = 2
  ) +
  ylab("number of predicted clades") +
  #Prettify
  theme_minimal(base_family = "Avenir Next", base_size = 8) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(linewidth = 0.1, colour = "lightgrey"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(family = "Avenir Next", size = 8, colour = "#252525"),
    axis.text.x = element_blank(),
    axis.text.y = element_text(family = "Avenir Next", size = 5, colour = "#252525"),
    strip.text = element_blank()
  ) +
  NULL
p.clade_counts

ggsave(
  p.clade_counts,
  filename = paste0(PARAM$folder.base, "results/clade_counts.overview_both.pdf"),
  width = 90,
  height = 85,
  unit = "mm",
  scale = 1,
  device = cairo_pdf
)
```

Plot per-habitat overview, combined for Archaea and Bacteria.

```{r, fig.width=8, fig.height=6}
use.h5 <- c(
  "all",
  "human gut (adult, healthy)",
  "human gut (non-industrialized)",
  "human gut (infant)",
  "pig gut",
  "bovine gut",
  "mammalian gut (other)",
  "rumen",
  "animal gut (other)",
  "human oral",
  "human skin",
  "human airways",
  "human urogenital",
  "built environment",
  "wastewater",
  "rhizosphere",
  "phyllosphere",
  "soil",
  "wetland",
  "marine",
  "fresh water",
  "groundwater",
  "hot spring",
  "hydrothermal vent",
  "aquatic (other)",
  "other"
)

#Consolidate habitat mapping data
p.clades.by_h5.all <- collect.h5_counts.bac %>%
  mutate(domain = "Bacteria") %>%
  #Summarise by habitat and (predicted) clade
  select(domain, marker_id, level, node, h5) %>%
  distinct() %>%
  left_join(collect.clade_sizes.bac, by = c("marker_id", "level", "node")) %>%
  #Count predicted clades per marker gene and habitat
  group_by(domain, marker_id, level, source, h5) %>%
  tally %>%
  #Merge archaeal data
  bind_rows(
    collect.h5_counts %>% filter(marker_id %in% use.marker_id.arch) %>% mutate(domain = "Archaea")
  ) %>%
  #Count prevalence of predicted clades across habitat
  left_join(tot.clade_sizes.by_source %>% ungroup(), by = c("domain", "marker_id", "source", "level")) %>%
  mutate(prevalence = n / n.total) %>%
  group_by(domain, level, h5, source) %>%
  #Normalize by total clade counts
  summarise(
    prevalence = mean(prevalence, na.rm=T),
    count = mean(prevalence, na.rm=T) * median(n.total, na.rm=T)
  ) %>%
  ungroup() %>%
  left_join(tot.clade_sizes.summarised.for_plotting, by = c("domain", "level", "source")) %>%
  mutate(count.adj = count + bl) %>%
  #Add global counts
  bind_rows(
    tot.clade_sizes.summarised.for_plotting %>%
      mutate(h5 = "all") %>%
      rename(count.adj = ceil)
  ) %>%
  #Re-adjust levels of relevant factors
  filter(!is.na(h5)) %>%
  filter(h5 %in% use.h5) %>%
  mutate(h5 = droplevels(fct_relevel(h5, c("all", levels(collect.h5_counts$h5))))) %>%
  mutate(level = fct_relevel(level, "phylum", "class", "order", "family", "genus")) %>%
  mutate(source = fct_relevel(source, rev(c("pg3", "gtdb_r220", "spire.mag", "spire.unbinned")))) %>%
  mutate(
    xmin = as.numeric(h5) - 0.5,
    xmax = as.numeric(h5) + 0.5
  ) %>%
  filter(level %in% c("phylum", "genus")) %>%
  #Plot
  ggplot(aes(x = h5, y = count.adj, ymin = bl, group = source, fill = source)) +
  #Add low-opacity background shading rectangles
  geom_rect(
    data = tot.clade_sizes.summarised.for_plotting %>%
      mutate(source = fct_relevel(source, rev(c("pg3", "gtdb_r220", "spire.mag", "spire.unbinned")))) %>%
      filter(level %in% c("phylum", "genus")),
    aes(x = length(use.h5), xmin = 0.5, xmax = length(use.h5) + 0.5, y = ceil, ymin = bl, ymax = ceil, fill = source),
    alpha = 0.1
  ) +
  #Add data series
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymax = count.adj),
    colour = "darkgrey",
    linewidth = 0.1
  ) +
  #Mark borders between sources
  geom_hline(
    data = tot.clade_sizes.summarised.for_plotting %>%
      mutate(source = fct_relevel(source, rev(c("pg3", "gtdb_r220", "spire.mag", "spire.unbinned")))) %>%
      filter(level %in% c("phylum", "genus")),
    aes(yintercept = ceil, colour = source),
    linetype = "dashed",
    linewidth = 0.1
  ) +
  #Adjust scales
  scale_fill_manual(values = rev(c(col.pg3, col.gtdb, col.spire.mag, col.unbinned.non_singleton))) +
  scale_colour_manual(values = rev(c(col.pg3, col.gtdb, col.spire.mag, col.unbinned.non_singleton))) +
  scale_x_discrete(guide = guide_axis(angle = -45), position = "top") +
  #scale_y_continuous(trans = "sqrt") +
  ylab("number of detected clade-level groups") +
  #Grid by taxonomic resolution
  facet_grid(domain + level ~ ., scales = "free_y") +
  #Prettify
  theme_minimal(base_family = "Avenir Next", base_size = 8) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(linewidth = 0.1, colour = "darkgrey"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(family = "Avenir Next", size = 8, colour = "#252525"),
    axis.text = element_text(family = "Avenir Next", size = 5, colour = "#252525"),
    strip.text = element_blank(),
  ) +
  NULL
p.clades.by_h5.all
  
ggsave(
  p.clades.by_h5.all,
  filename = paste0(PARAM$folder.base, "results/clade_counts.by_h5.all.pdf"),
  width = 90,
  height = 90,
  unit = "mm",
  device = cairo_pdf
)
```

# Export sample data table

```{r}
dat.sample.clean %>%
  filter(sample_id %in% dat.sample.rarefy$sample_id) %>%
  select(sample_id, ena_ers_sample_id, study_id, `age group:adolescent`:`terrestrial:wetland:salt marsh`) %>%
  left_join(dat.sample.rarefy %>% select(sample_id, h5), by = "sample_id") %>%
  rename(spire_sample_id = sample_id) %>%
  relocate(spire_sample_id, ena_ers_sample_id, study_id, h5) %>%
  pivot_longer(`age group:adolescent`:`terrestrial:wetland:salt marsh`) %>%
  filter(value) %>%
  mutate(
    name2 = str_remove(name, ".+:")
  ) %>%
  select(-name) %>%
  select(-value) %>%
  group_by(spire_sample_id, ena_ers_sample_id, study_id, h5) %>%
  summarise(microntology = str_c(name2, collapse = ";")) %>%
  ungroup() %>%
  arrange(study_id, spire_sample_id) %>%
  rename(habitat_category = h5) %>% write_tsv(file = "~/Desktop/table.S2.sample_metadata.tsv")

```




----


